
##challenges:
--> getting the legend working in quantiles or similar. just not percents. (check mark)
--> maybe getting multiple legends (check mark)
--> spatial join!



---
title: "Maryland Sports Venues Data Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(janitor)
library(broom)
library(ggplot2)
library(gt)
library(leaflet)
library(dplyr)
library(readxl)
library(shinydashboard)
library(rgdal)
library(htmltools)
library(leaflegend)
```


```{r}

library(dplyr)
library(readxl)
library(shinydashboard)
library(rgdal)
library(leaflet)
library(htmltools)
shapes <- rgdal::readOGR("cb_2018_us_county_5m","cb_2018_us_county_5m")

# the table starts from row 5
df_geo <- read_excel('all-geocodes-v2019.xlsx',skip=4)  %>% 
  filter(`Summary Level`=='040') %>%
  select(`State Code (FIPS)`, `Area Name (including legal/statistical area description)`)

colnames(df_geo) <- c('STATEFP','STATENAME')

shapes@data <- shapes@data %>% 
  left_join(df_geo) %>%
  mutate(ALAND = as.numeric(as.character(ALAND)),
         AWATER = as.numeric(as.character(AWATER)),
         NAME = as.character(NAME))
# remove shapes that are not in a state (e.g., Guam)
shapes <- shapes[!is.na(shapes@data$STATENAME),] 
# to be used by the dropdown list as available options
names_state <- sort(df_geo$STATENAME)

```
```{r}

md_shp <- st_read("https://raw.githubusercontent.com/frankrowe/maryland-geojson/master/maryland.geojson")

library(tidycensus)
census_api_key("0f4adff4206717d4947de6eb32d22e0b484ae6b8", install = TRUE, overwrite = TRUE)


md_income_tract <- get_acs(geography = "tract", 
                     variables = "B19013_001", 
                     state = "MD",
                     geometry = TRUE)

md_black_pop_tract <- get_acs(geography = "tract", 
              variables = "B02001_003", 
              state = "MD", 
              geometry = TRUE)

```

```{r}

pal5 <- colorQuantile(palette = "BuPu", domain = md_income_tract$estimate, n = 10)

pal6 <- colorQuantile(palette = "BuPu", domain = md_black_pop_tract$estimate, n = 10)

numPal <- colorNumeric('viridis', md_income_tract$estimate)

quantPal <- colorQuantile('viridis', md_income_tract$estimate, n = 5)

pal <- colorNumeric(
  palette = "viridis",
  domain = md_income_tract$estimate, na.color = NA)

mypal <- colorNumeric(
  palette = "viridis",
  domain = md_black_pop_tract$estimate, na.color = NA)

```

```{r}
md_map_all <- 
  leaflet()%>%
  addProviderTiles(providers$CartoDB.Positron)%>%
  addPolylines(
    data=md_shp,
    weight=2,
    color="gray"
  )%>%
  addPolygons(data = md_income_tract, group = "Income by Tract", color = ~ pal(estimate), stroke = FALSE)%>%
  addPolygons(data = md_black_pop_tract, group = "Black Population by Tract", color = ~ mypal(estimate), stroke = FALSE)%>%
  addLegend("bottomleft", pal = pal, values = md_income_tract$estimate, title = "Income Levels")%>%
  addLegend("bottomleft", pal = mypal, values = md_black_pop_tract$estimate, title = "Black Population")%>%
   addLayersControl(
  baseGroups = c("Income by Tract", "Black Population by Tract"),
  options = layersControlOptions(collapsed = FALSE)) 
  
md_map_all
  
```
#Scratch COde  
```{r}
  
  
  addPolygons(data = md_black_pop_tract, group = "Black Population by Tract", color = ~ pal6(estimate), stroke = FALSE)%>%
  addLegendNumeric(
    pal = numPal,
    values = md_income_tract$estimate,
    position = 'bottomleft')%>%
   addLayersControl(
  baseGroups = c("Income by Tract", "Black Population by Tract"),
  options = layersControlOptions(collapsed = FALSE))   
    
    
    
    "bottomleft", pal = pal5, title="Income Levels",
            values = md_income_tract$estimate, group="Income by Tract") 

```





%>%
  addLegendNumeric("bottomleft", pal = pal2, title="Black Population",
            values = md_black_pop_tract$estimate, group="Black Population by Tract") %>%
  addLayersControl(
  baseGroups = c("Income by Tract", "Black Population by Tract"),
  options = layersControlOptions(collapsed = FALSE))

md_map_all
```



```{r}
observeEvent(input$map_groups, {
  my_map <- leafletProxy("md_map_all") %>% clearControls()
  
  if (input$map_groups == 'Income by Tract'){
    my_map <- my_map %>%
      addLegend(
        "topright",
        pal = colorBin('OrRd', md_shp$md_income_tract),
        values = md_shp$md_income_tract)
  }else{
    my_map <- my_map %>%
      addLegend(
        "topright",
        pal = colorBin('YlGnBu', md_shp$md_black_pop_tract),
        values = md_shp$md_black_pop_tract)
  }
})
  
```




