```{r}
library(tidyverse)
library(postmastr)
library(usdata)
library(ggplot2)
library(scales)
library(stringr)
```

```{r}
wbb_rosters23 <- read_csv("https://github.com/dwillis/wbb-rosters/raw/main/rosters_2022-23.csv")
```

# Teams That Have Roster Data

```{r}
teams <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/teams.csv")

roster_teams <- wbb_rosters23 %>% distinct(ncaa_id) %>% mutate(has_roster = 1)

teams <- teams %>% 
  left_join(roster_teams) %>% mutate(roster = has_roster)

teams_without_rosters <- teams %>% 
  filter(is.na(roster))
```


# Things to Clean

* Drop where Name is NA
* Standardize Positions
* Standardizing State (would need to extract)
* Standardizing Year (dealing with redshirts & also grad students)
* Standardizing Height (5'10" or 5-10)

# Columns to Add

* State
* Country
* Height in inches

# Height Cleaning

* Still need to add a single column with total inches using feet and inches.

```{r}
wbb_rosters23 <- wbb_rosters23 %>% mutate(
  height = str_replace(height, '-', "'")
)
wbb_rosters23 <- wbb_rosters23 %>% separate(col=height, into=c("height_ft", "height_in"), sep="'", extra="merge")
wbb_rosters23 <- wbb_rosters23 %>% mutate(height_in = str_replace(height_in, '"', " "))
wbb_rosters23$height <- paste(wbb_rosters23$height_ft, "-", wbb_rosters23$height_in)
```


# Positions Cleaning

```{r}

positions_cleaned <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/wbb_positions.csv") %>%
  select(-count)

wbb_rosters23 <- wbb_rosters23 %>% inner_join(positions_cleaned, by = "position") 

wbb_rosters23 %>% distinct(position_clean)

## Primary/Secondary positions based on order of position_clean

wbb_rosters23 <- wbb_rosters23 %>% mutate(position_full = position_clean)
wbb_rosters23 <- wbb_rosters23 %>% mutate(position_clean = na_if(position_clean, "N/A"))
wbb_rosters23 <- wbb_rosters23 %>% separate(position_clean, c('primary_position', 'secondary_position'), sep='/', extra='merge')
wbb_rosters23 <- wbb_rosters23 %>% mutate(primary_position = case_when((primary_position == "G") ~ "Guard", (primary_position == "C") ~ "Center", (primary_position == "F") ~ "Forward", TRUE ~ primary_position)) 
wbb_rosters23 <- wbb_rosters23 %>% mutate(secondary_position = case_when((secondary_position == "G") ~ "Guard", (secondary_position == "C") ~ "Center", (secondary_position == "F") ~ "Forward", TRUE ~ secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% mutate(primary_position = case_when((primary_position == "G") ~ "Guard", (primary_position == "C") ~ "Center", (primary_position == "F") ~ "Forward", TRUE ~ primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("'", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("5", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("6", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("'", wbb_rosters23$secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("5", wbb_rosters23$secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("6", wbb_rosters23$secondary_position))
wbb_rosters23 %>% group_by(primary_position) %>% summarise(count = n()) %>% arrange(desc(count))
wbb_rosters23 %>% group_by(secondary_position) %>% summarise(count = n()) %>% arrange(desc(count))
  
```

# Years Cleaning

```{r}
years_cleaned <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/yearscleaned.csv")

wbb_rosters23 <- inner_join(wbb_rosters23, years_cleaned) %>% rename(year_clean = `year-clean`)

wbb_rosters23 <- wbb_rosters23 %>% 
  select(-count)

wbb_rosters23 %>% distinct(year_clean)
```

# Previous School Cleaning

Some records list the player's high school in the previous school column. We filter those out and turn it into a list, then update `wbb_rosters23` to replace hs_clean with the previous school for those rows.

```{r}
hs_wbb <- wbb_rosters23 %>%
  filter(is.na(high_school))%>%
  filter(!is.na(previous_school))%>%
  filter(str_detect(previous_school, "HS"))%>%
  mutate(hs_clean = "")%>%
  mutate(hs_clean = previous_school)%>%
  mutate(type = "has hs")
extras_wbb <- wbb_rosters23 %>%
  filter(!is.na(high_school))%>%
  filter(is.na(previous_school))%>%
  mutate(hs_clean = "")%>%
  mutate(type = "no hs")
all_wbb <- bind_rows(hs_wbb, extras_wbb)

previous_school_is_hs=as.list(hs_wbb$previous_school)

# update based on list membership

wbb_rosters23 <- wbb_rosters23 %>%
  mutate(hs_clean = case_when(
    is.na(high_school) & previous_school %in% previous_school_is_hs ~ previous_school,
    TRUE ~ high_school)
  )

# where previous school == hs_clean, replace previous_school with NA

wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(previous_school = ifelse(previous_school == hs_clean, NA, previous_school))
```

# Hometown/Homestate Cleaning

```{r}
wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(hometown_cleaned = str_to_upper(str_replace(hometown, "\\.",""))) %>% 
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"\\.","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"/.*","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"\\.$","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,",",", ")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"-"," ")) %>% 
  mutate(hometown_cleaned = str_squish(hometown_cleaned))

states_non_standard <- pm_append(type = "state", input = c("SD.", "MASS", "CALIF", "MICH", "NEB", "IND", "MINN", "ORE", "OHIO", "FLA", "MISS", "TENN", "ARIZ", "KAN", "ALA", "OKLA", "WIS", "ILL", " WASH", "ARK", "COLO", "NEV", "CONN", "WISC", "WVA", "DEL", "WYO", "CALI", "LOUIS", "VIRG", "MONT", "PENN", "TEX", "KANS", "NEBR", "IDA", "COL"), output = c("SD", "MA", "CA", "MI", "NE", "IN", "MN", "OR", "OH", "FL", "MS", "TN", "AZ", "KS", "AL", "OK", "WI", "IL", "WA", "AR", "CO", "NV", "CT", "WI", "WV", "DE", "WY", "CA", "LA", "VA", "MT", "PA", "TX", "KS", "NE", "ID", "CO"), locale = "us")

dict <- pm_dictionary(type='state', case = c("title", "upper", "lower"), append = states_non_standard)

wbb_rosters23 <- wbb_rosters23 %>% pm_identify(var="hometown_cleaned")

parsed <- wbb_rosters23 %>% 
  pm_identify(var="hometown_cleaned") %>% 
  pm_prep(var="hometown_cleaned", type="street") %>% 
  pm_state_parse(dict)

wbb_rosters23 <- left_join(wbb_rosters23, parsed, by="pm.uid")

wbb_rosters23 %>%
  filter(is.na(hometown_cleaned))

wbb_rosters23 <- wbb_rosters23 %>% separate(hometown, c('hometown', 'homestate'), sep=',', extra='merge')

wbb_rosters23 <- wbb_rosters23 %>% mutate(homestate=str_trim(homestate), homestate=str_replace(homestate, '\\.', '')) %>%
  mutate(homestate = case_when(is.na(abbr2state(homestate)) ~ homestate, TRUE ~ (abbr2state(homestate))))

```

# Foreign Nation Cleaning

Take the list of FIBA nations (https://www.fiba.basketball/rankingwomen) and make a list or dataframe of them, then compare that to the values in the `homestate` column generated above and populate a new column for country.

```{r}

```



# Generate summary stats for pm.state, year_clean and position_full and make graphics (Rina & Mike)

```{r}
states <- wbb_rosters23%>%
  drop_na(pm.state)%>%
  group_by(pm.state)%>%
  summarise(count = n())%>%
  arrange(desc(count))%>%
  top_n(10)


ggplot() + 
  geom_bar(data=states, 
  aes(x=reorder(pm.state, count), weight=count, fill = "red")) +
  coord_flip()+
  labs(title="Top ten states where women's college basketball players come from", subtitle="More than 800 college women's basketball players hail from Pennsylvania.", caption="Source: NCAA", x = "State", y= "Count of players") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank(),
    legend.position = "none"
    )
  
```


```{r}

year <- wbb_rosters23%>%
  group_by(year_clean)%>%
  summarise(count = n())%>%
  arrange(desc(count))

ggplot() + 
  geom_bar(data=year, 
  aes(x=reorder(year_clean, -count), weight=count, fill="red"))+
  scale_y_continuous(labels=comma)+
  labs(title="More than 3,000 college women's basketball players are freshmen", caption="Source: NCAA", x = "Position", y= "Count of players") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 8), 
    panel.grid.minor = element_blank(),
    legend.position = "none"
    )

```

```{r}



positions <- wbb_rosters23%>%
  group_by(position_full)%>%
  summarise(count = n())%>%
  arrange(desc(count))


positions$position_full = str_to_title(positions$position_full)

ggplot() + 
  geom_bar(data=positions, 
  aes(x=reorder(position_full, count), weight=count, fill="red"))+
  scale_y_continuous(labels=comma)+
  labs(title="The bulk of college women's basketball players are guards", caption="Source: NCAA", x = "Position", y= "Count of players") + 
  coord_flip()+
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 8), 
    panel.grid.minor = element_blank(),
    legend.position = "none"
    )

````

