```{r}
library(tidyverse)
library(postmastr)
library(usdata)
library(rvest)
```

```{r}
wbb_rosters23 <- read_csv("https://github.com/dwillis/wbb-rosters/raw/main/rosters_2022-23.csv")
```

# Teams That Have Roster Data

```{r}
teams <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/teams.csv") %>% select(-url, -twitter, -stats_name, -private)

roster_teams <- wbb_rosters23 %>% distinct(ncaa_id) %>% mutate(has_roster = 1)

teams <- teams %>% 
  left_join(roster_teams) %>% mutate(roster = has_roster)

teams_without_rosters <- teams %>% 
  filter(is.na(roster))

wbb_rosters23 <- wbb_rosters23 %>% 
  left_join(teams, by=c("ncaa_id", "team")) 
```

# Height Cleaning
# Populate heights for any team with more than 5 players missing them - see Height Cleaning section for list (Rina)

```{r}
wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(height_clean = str_replace(height, "'", '-')) %>% 
  mutate(height_clean = str_replace(height_clean, "’", '-')) %>% 
  mutate(height_clean = str_replace(height_clean, "''", '')) %>% 
  mutate(height_clean = str_replace(height_clean, '"', '')) %>%
  separate(height_clean, c('height_ft', 'height_in'), sep="-", extra="merge") %>% 
  mutate(height_ft = as.numeric(height_ft), height_in = as.numeric(height_in)) %>%
  mutate(total_inches = (height_ft*12)+height_in)


wbb_rosters23 %>% 
  group_by(total_inches) %>% 
  summarise(
    total = n()
  )
```

We've got 98 with NA heights.

```{r}
#Identify teams with missing heights
wbb_rosters23 %>% filter(is.na(total_inches)) %>% group_by(team) %>% summarize(count = n()) %>% arrange(desc(count))%>%
  filter(count>5)

#Create list of teams with more than five players missing heights
five_plus_list <- c("Geneva", "Salem (WV)", "Pomona-Pitzer", "Oakland")

#Create dataframe of all players on teams with more than five players missing heights
five_plus_filtered <- wbb_rosters23 %>%
  filter(team %in% five_plus_list)

#Write out to CSV
write.csv(five_plus_filtered,"jour479x_fall_2022/assignments/height_cleaning.csv", row.names = FALSE)


```

#To populate missing heights, I manually inputted data from rosters I found online on team websites. If heights were not available online, in the case of Oakland, I pulled data from ESPN. Here are the sites I looked at.
--> https://salemtigers.com/sports/womens-basketball/roster
--> https://athletics.geneva.edu/sports/womens-basketball/roster/taylor-haring/2887 
--> https://goldengrizzlies.com/sports/womens-basketball/roster
--> https://www.espn.com/womens-college-basketball/team/roster/_/id/2473/oakland-golden-grizzlies 
--> https://www.sagehens.com/sports/wbkb/2021-22/roster 

#After manually inputting the heights into a csv, I brought the dataframe back into R.

```{r}

cleaned_heights <- read.csv("height_cleaning.csv")

cleaned_heights

```


##Here's some scratch code attempts on how we might join them back.
```{r}
wbb_rosters23_minus <- wbb_rosters23%>%
  filter(!team %in% five_plus_list)

clean_heights_join <- inner_join(cleaned_heights, five_plus_filtered, by="player_id")

rejoin <- bind_rows(wbb_rosters23_minus, clean_heights_join)

```


#I identified several issues with the data during this process.
--> Mariona Garcia of Salem (WV) was not on team's official roster.
--> Neither ESPN nor the official Oakland roster listed height for Brooke Daniels.
--> Neither ESPN nor the official Oakland roster listed height for Eve Pardo.
--> Neither ESPN nor the official Oakland roster listed height for Lianna Baxter of Oakland.
--> Pomona-Pitzer is a special case and requires a different scrape become of the nature of the joint nature of the team. Here is the link to the roster: https://www.sagehens.com/sports/wbkb/2022-23/roster 

#View the Pomona-Pitzer data below.
Names and player_ids were not populated. Positions appear in the name column. Height appears in the year column.
```{r}
wbb_rosters23 %>% filter(is.na(total_inches)) %>% 
  filter(team == "Pomona-Pitzer")

```


# Positions Cleaning

```{r}

positions_cleaned <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/wbb_positions.csv") %>%
  select(-count)

wbb_rosters23 <- wbb_rosters23 %>% inner_join(positions_cleaned, by = "position") 

wbb_rosters23 %>% distinct(position_clean)

## Primary/Secondary positions based on order of position_clean

wbb_rosters23 <- wbb_rosters23 %>% mutate(position_full = position_clean)
wbb_rosters23 <- wbb_rosters23 %>% mutate(position_clean = na_if(position_clean, "N/A"))
wbb_rosters23 <- wbb_rosters23 %>% separate(position_clean, c('primary_position', 'secondary_position'), sep='/', extra='merge')
wbb_rosters23 <- wbb_rosters23 %>% mutate(primary_position = case_when((primary_position == "G") ~ "Guard", (primary_position == "C") ~ "Center", (primary_position == "F") ~ "Forward", TRUE ~ primary_position)) 
wbb_rosters23 <- wbb_rosters23 %>% mutate(secondary_position = case_when((secondary_position == "G") ~ "Guard", (secondary_position == "C") ~ "Center", (secondary_position == "F") ~ "Forward", TRUE ~ secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% mutate(primary_position = case_when((primary_position == "G") ~ "Guard", (primary_position == "C") ~ "Center", (primary_position == "F") ~ "Forward", TRUE ~ primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("'", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("5", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("6", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("'", wbb_rosters23$secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("5", wbb_rosters23$secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("6", wbb_rosters23$secondary_position))
wbb_rosters23 %>% group_by(primary_position) %>% summarise(count = n()) %>% arrange(desc(count))
wbb_rosters23 %>% group_by(secondary_position) %>% summarise(count = n()) %>% arrange(desc(count))
  
```

# Years Cleaning

```{r}
years_cleaned <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/yearscleaned.csv")

wbb_rosters23 <- inner_join(wbb_rosters23, years_cleaned) %>% rename(year_clean = `year-clean`)

wbb_rosters23 <- wbb_rosters23 %>% 
  select(-count)

wbb_rosters23 %>% distinct(year_clean)
```

# Previous School Cleaning

Some records list the player's high school in the previous school column. We filter those out and turn it into a list, then update `wbb_rosters23` to replace hs_clean with the previous school for those rows.

```{r}
hs_wbb <- wbb_rosters23 %>%
  filter(is.na(high_school))%>%
  filter(!is.na(previous_school))%>%
  filter(str_detect(previous_school, "HS"))%>%
  mutate(hs_clean = "")%>%
  mutate(hs_clean = previous_school)%>%
  mutate(type = "has hs")
extras_wbb <- wbb_rosters23 %>%
  filter(!is.na(high_school))%>%
  filter(is.na(previous_school))%>%
  mutate(hs_clean = "")%>%
  mutate(type = "no hs")
all_wbb <- bind_rows(hs_wbb, extras_wbb)

previous_school_is_hs=as.list(hs_wbb$previous_school)

# update based on list membership

wbb_rosters23 <- wbb_rosters23 %>%
  mutate(hs_clean = case_when(
    is.na(high_school) & previous_school %in% previous_school_is_hs ~ previous_school,
    TRUE ~ high_school)
  )

# where previous school == hs_clean, replace previous_school with NA

wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(previous_school = ifelse(previous_school == hs_clean, NA, previous_school))
```

# Hometown/Homestate Cleaning

```{r}
wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(hometown_cleaned = str_to_upper(str_replace(hometown, "\\.",""))) %>% 
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"\\.","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"/.*","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"\\.$","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,",",", ")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"-"," ")) %>% 
  mutate(hometown_cleaned = str_squish(hometown_cleaned))

states_non_standard <- pm_append(type = "state", input = c("SD.", "MASS", "CALIF", "MICH", "NEB", "IND", "MINN", "ORE", "OHIO", "FLA", "MISS", "TENN", "ARIZ", "KAN", "ALA", "OKLA", "WIS", "ILL", " WASH", "ARK", "COLO", "NEV", "CONN", "WISC", "WVA", "DEL", "WYO", "CALI", "LOUIS", "VIRG", "MONT", "PENN", "TEX", "KANS", "NEBR", "IDA", "COL"), output = c("SD", "MA", "CA", "MI", "NE", "IN", "MN", "OR", "OH", "FL", "MS", "TN", "AZ", "KS", "AL", "OK", "WI", "IL", "WA", "AR", "CO", "NV", "CT", "WI", "WV", "DE", "WY", "CA", "LA", "VA", "MT", "PA", "TX", "KS", "NE", "ID", "CO"), locale = "us")

dict <- pm_dictionary(type='state', case = c("title", "upper", "lower"), append = states_non_standard)

wbb_rosters23 <- wbb_rosters23 %>% pm_identify(var="hometown_cleaned")

parsed <- wbb_rosters23 %>% 
  pm_identify(var="hometown_cleaned") %>% 
  pm_prep(var="hometown_cleaned", type="street") %>% 
  pm_state_parse(dict)

wbb_rosters23 <- left_join(wbb_rosters23, parsed, by="pm.uid")

wbb_rosters23 %>%
  filter(is.na(hometown_cleaned))

wbb_rosters23 <- wbb_rosters23 %>% separate(hometown, c('hometown', 'homestate'), sep=',', extra='merge')

wbb_rosters23 <- wbb_rosters23 %>% mutate(homestate=str_trim(homestate), homestate=str_replace(homestate, '\\.', '')) %>%
  mutate(homestate = case_when(is.na(abbr2state(homestate)) ~ homestate, TRUE ~ (abbr2state(homestate))))

```

# Foreign Nation Cleaning

Take the list of FIBA nations (https://www.fiba.basketball/rankingwomen) and make a list or dataframe of them, then compare that to the values in the `homestate` column generated above and populate a new column for country.

```{r}
fibaurl <- "https://www.fiba.basketball/rankingwomen"
nations <- fibaurl %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="fiba_ranking_table_wrapper"]/table') %>%
  html_table()
nations_df <- nations[[1]]
wbb_rosters23 <- mutate(wbb_rosters23, country = case_when(!is.na(pm.state) ~ "USA"))


wbb_rosters23 <- wbb_rosters23 %>% mutate(temp = sapply(strsplit(hometown_cleaned, ", ", fixed=TRUE), tail, 1))

wbb_rosters23 <- mutate(wbb_rosters23, country = case_when(temp %in% c('BC', 'QUEBEC', 'BRITISH COLUMBIA', 'ALBERTA', 'ONTARIO') ~ "CANADA", TRUE ~ country))

nations_df$Country <- toupper(nations_df$Country)



added_nations <- data.frame(Worldrank=c(0, 0, 0, 0),
                  Country=c('ENGLAND', 'RUSSIA', 'SCOTLAND', 'NORTHERN IRELAND'),
                  Zonerank=c(0, 0, 0, 0),
                  IOC=c('', '', '', ''),
                  "Current points"=c(0, 0, 0, 0),
                  "+/- Rank *"=c(0, 0, 0, 0),
                  check.names = FALSE)
nations_df <- rbind(nations_df, added_nations)


wbb_rosters23 <- left_join(wbb_rosters23, nations_df, by=c('temp'='Country'))

wbb_rosters23 <- mutate(wbb_rosters23, country = case_when((!is.na(Worldrank) & is.na(country)) ~ temp, is.na(Worldrank) &  !is.na(country) ~ country, country == "USA" ~ "USA"))

wbb_rosters23 <- subset(wbb_rosters23, select = -c(Worldrank,Zonerank,IOC,`Current points`, `+/- Rank *`) )

wbb_rosters23 %>% filter(is.na(country)) %>% group_by(temp) %>% summarise(count=n()) %>% arrange(desc(count))
```




