---
title: "Women's Soccer Data"
output: html_notebook
---

We'll use this notebook to explore some NCAA women's soccer data and [our questions](https://docs.google.com/spreadsheets/d/1kTSrnl1URmcexIWLDMWqDnp3q-c2OS45uifnhjp10n4/edit#gid=0). First, let's load the tidyverse library:

```{r}
library(tidyverse)
```

We'll want to start with 2022 match data, which is located in [this GitHub repository](https://github.com/dwillis/NCAAWomensSoccerData). We can load the match data directly from GitHub using `read_csv`:


```{r}
matches_2022 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2022.csv")
```

Let's take a look at this data to see what it contains, using `head` and `glimpse`:

```{r}
head(matches_2022)
glimpse(matches_2022)
```

Now, onto the questions using aggregates, which means `group_by` and `summarize`:

1. How many shots in total has Maryland attempted and how many have been on goal? Calculate a percentage of shots on goal.

```{r}
matches_2022 %>% 
  filter(team == 	'Maryland Terrapins, Terps')%>%
  #group_by(team)%>%
  summarise(total_shots = sum(sh_att), total_quality_shots = sum(so_g))%>%
  mutate(percentage_quality_shots = (total_quality_shots/total_shots)*100)%>%
  arrange(desc(percentage_quality_shots))
```

2. What about the same shot stats from last season's first five games? You'll need to load the matches from 2021 - just change the season.

```{r}
matches_2021 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2021.csv")



```

```{r}

matches_2021 %>%
  filter(date < '2021-09-03')%>%
  filter(team == 	'Maryland Terrapins, Terps')%>%
  summarise(total_shots = sum(sh_att), total_quality_shots = sum(so_g))%>%
  mutate(percentage_quality_shots = (total_quality_shots/total_shots)*100)%>%
  arrange(desc(percentage_quality_shots))

  
```


3. What about draws - Maryland started with 4 ties. Have any other teams had as many?

```{r}
matches_2022 %>%
  group_by(team)%>%
  filter(result == 'T')%>%
  summarise(tie_count =n())%>%
  arrange(desc(tie_count))
  

  
```


4. How many total corner kicks has Maryland had so far and how many have its opponents have? Calculate the difference between the two totals for every team.

```{r}
#total no. of terps corners
terps_corners <- matches_2022 %>% 
  filter(team == 	'Maryland Terrapins, Terps')%>%
  summarise(terps_corners = sum(corners))

#calculate corners v. opp corners, pull in terps corners for ref
matches_2022 %>% 
  group_by(team)%>%
  summarise(terps_corners, total_corners = sum(corners), opp_corners = sum(defensive_corners))%>%
  mutate(difference = (total_corners - opp_corners))


```
```{r}
#difference from terps total corners

matches_2022 %>% 
  group_by(team)%>%
  summarise(total_corners = sum(corners), terps_corners)%>%
  mutate(difference_from_terps_corners = (terps_corners - total_corners))%>%
  arrange(desc(difference_from_terps_corners))


```

