---
title: "background"
output: html_document
date: "2022-10-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Working Question: How has the gap between men's and women's sports changed at UMD over the years?
This question is multifaceted. Through equity reports, I plan to explore the following elements over the past ten years:
--> The difference in revenue for men's and women's sports
--> The difference in total expenses (including game-day expenses) for men's and women's sports
--> The difference in expenses for recruiting in men's and women's sports
--> Bonus: attendance data, would require a PIA request

#About the data
The data comes from the Equity in Athletics Data Analysis report, which is filed by co-ed postsecondary institutions that receive Title IV funding and have an intercollegiate athletics program. The U.S. Department of Education's Office of Postsecondary Education has a data tool that allows users to download elements of the report. The data tool includes data on the University of Maryland from 2003 to 2020. I also identified a document from the University System of Maryland that includes data published in 2022 for the EADA report. Also of note, the data tool seems to make comparing institutions simple. It could be interesting to compare data across the Big 10 over the past decade.

--> Here is the data tool: https://ope.ed.gov/athletics/#/institution/details 
--> Here is the USM data: https://www.usmd.edu/regents/agendas/20220613-ICA-PublicSession.pdf 

#Cleaning and preparing the data
The data that I've identified will likely need to be combined into one dataframe. I could run the USM data through Tabula (though I expect it to messy) and add it to the data extracted from the data tool. It appears the historical part of the data tool can extract data from 2003 to 2020, but another part includes data from 2021, and the USM data includes data from 2022. I'll need to seek some clarity here and adjust and arrange the data accordingly.

#A look at some of the data extracted from the data tool.
```{r}

#load tidyverse
library(tidyverse)

#read in datasets
expenses <- read_csv('expenses.csv')
op_expenses <- read_csv('operating_expenses.csv')
recruiting <- read_csv('recruiting_expenses.csv')
revenue <- read_csv('revenues.csv')

#remove columns that are entirely NAs (that includes sports like Archery, for example, which doesn't have program at UMD)
expenses <- expenses[ , colSums(is.na(expenses)) < nrow(expenses)]
op_expenses <- op_expenses[ , colSums(is.na(op_expenses)) < nrow(op_expenses)]
recruiting <- recruiting[ , colSums(is.na(recruiting)) < nrow(recruiting)]
recruiting$`Institution Name` <- tolower(recruiting$`Institution Name`)
revenue <- revenue[ , colSums(is.na(revenue)) < nrow(revenue)]

#take a look
head(expenses)
head(op_expenses)
head(recruiting)
head(revenue)

```

#Taking a closer look at revenue, and a visualization attempt
```{r}

#load ggplot2 and lubridate
library(ggplot2)
library(lubridate)

#to isolate
total_revenue <-
  revenue %>%
  select("Survey Year","Total Men's Team Revenue", "Total Women's Team Revenue")

#pivot to prep for visualization
total_revenue <-
total_revenue %>% pivot_longer(cols=c("Total Men's Team Revenue", "Total Women's Team Revenue"),
                    names_to='team',
                    values_to='revenue')

#a line chart attempt
ggplot(data=total_revenue, aes(x=`Survey Year`, y=revenue, group = team, color = team)) +
  geom_line()+
  geom_point() +
 labs(
    title="Maryland is earning more on men's sports than on women's sports", 
    x="Year", 
    y="Dollar amount") + 
  scale_y_continuous(labels = scales::comma, limits = c(0, 80000000)
  )

```


#Regarding the bonus data
I also plan to request historic attendance data, whether it fits directly into the purview of this project or not. I plan to send a PIA request to the University of Maryland requesting data on attendance for all sports over the past ten years by game. I expect this might be an expensive request, but it's worth a shot. 

Here's where I will send the PIA: https://www.umd.edu/administration/public-information-request 

#Looking ahead, questions and concerns
My current concern is that I might be heading in a bit too broadly, but I think I can continue to use the bullet points I identified at the top to guide my reporting. I'm not sure how much data I'll really need here to tell a compelling story, but I suppose I can use the other bullet points to fall back on if I don't have a compelling story come out of the existing metrics. Still, I'm confident there's something I can pull from the report.

On another note, I'm also still thinking about the planned time frame to explore. I'm thinking about the past ten years, since I've been told ten years can be a good baseline for this sort of analysis, but I'm also open to looking at this data over a longer time frame. I'm not sure what the best course of action is here.

#Action items
Next up, on my things to do:
--> Send the PIA on attendance!
--> Get to the bottom of the data situation. Account for any skipped years. Compile the dataframe into something neater. Be sure to understand all variables fully.


#Other relevant works
This law firm conducted an external equity report on the NCAA. It can be viewed here: https://ncaagenderequityreview.com/ 

Here's an NPR report based on the external equity report: https://www.npr.org/2021/08/03/1024481199/report-ncaa-undervalues-womens-basketball-prioritizes-mens-teams 

Here's an AP story that shows that the NCAA spends more on male athletes than female athletes: https://www.espn.com/college-sports/story/_/id/32481915/second-ncaa-gender-equity-report-shows-more-money-spent-male-athletes-female-ones 

Here's an AP story about pay equality for NCAA referees: https://apnews.com/article/sports-college-basketball-march-madness-eabc98c307c975c7b56ce843a69d885f 

#Describing the data

#About expenses, op_expenses, recruiting and revenue dataframes
Time frame: 2003 - 2020
Row count: 15 
Row description: Each row is one year
Unit: Dollars
Details: The expenses dataframe includes dollars spent for each team. The op_expenses dataframe includes dollars spent on operating, or game day expenses. According to the EADA, this includes money spent on lodging, meals, transportation, uniforms, equipment, team members, support staff and more." The recruiting dataframe shows money spent on recruiting divided by men's and women's team. The revenue dataframe shows how much each team earned. Recruiting expenses and operating expenses are factored into the total expenses dataframe.

Data Dictionary Source: https://surveys.ope.ed.gov/athletics2k20/wwwroot/documents/2019_EADA_Users_Guide.pdf 

#Additional notes on the time frames:
The survey data ends in 2020, which represents data collected from 07/01/2020 - 06/30/2021. The University System of Maryland shared this data in its Board of Regents meetings in June of 2022. The good news is that I anticipated a need to join old and new data. The bad news is that this data is the most recent, but it's ageing quickly.

#A quick visualization of expense data over time
```{r}
library(tidyverse)
#expenses

expenses_mens <- expenses %>%
  select(`Survey Year`, `Total Men's Team Expenses`, `Total Women's Team Expenses`)

expenses_womens <- expenses %>%
  select(`Survey Year`, `Total Women's Team Expenses`)

ggplot() + 
  geom_line(data=expenses_mens, aes(x=`Survey Year`, y=`Total Men's Team Expenses`, color="Total Men's Team Expenses")) + 
  geom_line(data=expenses_womens, aes(x=`Survey Year`, y=`Total Women's Team Expenses`, color="Total Women's Team Expenses")) +
  labs(
    title="Maryland is spending much more on men's sports than on women's sports", 
    x="Year", 
    y="Dollar amount") + 
  scale_y_continuous(labels = scales::comma, limits = c(0, 50000000)
)


```

#Another visualization on recruiting expenses over time
```{r}

#recruiting expenses

rec_mens <- recruiting %>%
  select(`Survey Year`, `Men's Team Recruiting Expenses`)

rec_womens <- recruiting %>%
  select(`Survey Year`, `Women's Team Recruiting Expenses`)

ggplot() + 
  geom_line(data=rec_mens, aes(x=`Survey Year`, y=`Men's Team Recruiting Expenses`, color="Men's Team Recruiting Expenses")) + 
  geom_line(data=rec_womens, aes(x=`Survey Year`, y=`Women's Team Recruiting Expenses`, color="Women's Team Recruiting Expenses")) +
  labs(
    title="Maryland pours more money into men's sports recruiting than women's sports", 
    x="Year", 
    y="Dollar amount") + 
  scale_y_continuous(labels = scales::comma, limits = c(0, 1200000)
)
```

#Observations and next steps
Next, I want to take a look a look at the breakdown of where these expenses are coming from. I'm curious to see visualize which teams are spending what and which teams are spending the most. I'm also curious to see similar data from recruiting. It will be interesting to compare recruiting expenses for women's and men's basketball, for example, which should be more comparable. Men's football is likely to be playing a major role in these figures, so it will be interesting to inspect just how big that role is, and what the gap between men's and women's sports look like without football expenses. I also am eager to visualize and analyze data on revenues, similarly factoring in and factoring out football. I also plan to compare this against at least one other Big Ten schools.

#Here's a quick look at what incorporating the Big Ten might look like.

```{r}
#This is expense data on all Big Ten teams. It could be interesting to measure the size of the gap between each of these teams, as well.
big10_expenses <- read_csv('big10_expenses.csv')
big10_expenses$`Institution Name` <- tolower(big10_expenses$`Institution Name`)
big10_expenses$`Institution Name` <- gsub("rutgers university-new brunswick/piscataway", "rutgers", big10_expenses$`Institution Name`)
big10_expenses$`Institution Name` <- gsub("rutgers university-new brunswick", "rutgers", big10_expenses$`Institution Name`)
big10_expenses$`Institution Name` <- gsub("rutgers university-new brunswick/piscataway", "rutgers", big10_expenses$`Institution Name`)
big10_expenses$`Institution Name` <- gsub("university of illinois urbana-champaign", "illinois", big10_expenses$`Institution Name`)
big10_expenses$`Institution Name` <- gsub("university of illinois at urbana-champaign", "illinois", big10_expenses$`Institution Name`)


big10_selects <- big10_expenses %>%
  select(`Survey Year`, `Institution Name`,`Total Men's Team Expenses`, `Total Women's Team Expenses`)


mens <- ggplot() + 
  geom_line(data=big10_selects, aes(x=`Survey Year`, y=`Total Men's Team Expenses`,  group=`Institution Name`, color=`Institution Name`)) + 
  scale_y_continuous(labels = scales::comma, limits = c(0, 95000000)
)

mens

womens <- ggplot() + 
  geom_line(data=big10_selects, aes(x=`Survey Year`, y=`Total Women's Team Expenses`,  group=`Institution Name`, color=`Institution Name`)) + 
  scale_y_continuous(labels = scales::comma, limits = c(0, 95000000)
)

womens

#an attempt to plot these charts next to each other
#library(cowplot)
#plot_grid(mens, womens)

#Of course, the legend labels need to be fixed!

```
#Measuring the gap between expenses each year, some free experimentation
```{r}
big10_selects <- big10_selects%>%
  mutate(gap = `Total Men's Team Expenses` - `Total Women's Team Expenses` )

terps_expense_gap <- big10_selects%>%
  filter(`Institution Name` == "university of maryland-college park")

ggplot() + 
  geom_line(data=big10_selects, aes(x=`Survey Year`, y=`gap`, group=`Institution Name`), color="grey") + 
  geom_line(data=terps_expense_gap, aes(x=`Survey Year`, y=`gap`), color="red") + 
  labs(
    title="In the Big 10, the gap between men's and women's sports is smaller at UMD",
    y="expenses gap")+ 
  scale_y_continuous(labels = scales::comma, limits = c(0, 70000000)
)

#It looks like the Maryland sports gender gap is smaller than that of other Big Ten teams.

```

Are there outliers or patterns in recruiting spending? Do those coincide with success or coaching changes or performance?


#Recruting spending by sport, Maryland
```{r}

#clean up the data
expenses$`Institution Name` <- tolower(expenses$`Institution Name`)

#select teams of interest
md_teams_expenses <- expenses%>%
  select(`Institution Name`, `Survey Year`, `Basketball Men's Team Expenses`, `Basketball Women's Team Expenses`, `Baseball Total Expenses`,`Field Hockey Total Expenses`, `Football Total Expenses`, `Golf Men's Team Expenses`, `Golf Women's Team Expenses`, `Gymnastics Women's Team Expenses`, `Lacrosse Men's Team Expenses`, `Lacrosse Women's Team Expenses`, `Soccer Men's Team Expenses`, `Soccer Women's Team Expenses`, `Softball Total Expenses`, `Swimming and Diving Men's Team Expenses`, `Swimming and Diving Women's Team Expenses`, `Tennis Men's Team Expenses`, `Tennis Women's Team Expenses`,`Volleyball Total Expenses`,`Water Polo Total Expenses`, `Wrestling Total Expenses`)

#prep for visualization y pivoting longer
md_teams_expenses_long <-
md_teams_expenses %>% pivot_longer(cols=c("Basketball Men's Team Expenses", "Basketball Women's Team Expenses", "Baseball Total Expenses","Field Hockey Total Expenses", "Football Total Expenses", "Golf Men's Team Expenses", "Golf Women's Team Expenses", "Gymnastics Women's Team Expenses", "Lacrosse Men's Team Expenses", "Lacrosse Women's Team Expenses", "Soccer Men's Team Expenses", "Soccer Women's Team Expenses", "Softball Total Expenses", "Swimming and Diving Men's Team Expenses", "Swimming and Diving Women's Team Expenses", "Tennis Men's Team Expenses", "Tennis Women's Team Expenses","Volleyball Total Expenses","Water Polo Total Expenses", "Wrestling Total Expenses"),
                    names_to='team',
                    values_to='expenses')
#let's visualize
ggplot() + 
  geom_line(data=md_teams_expenses_long, aes(x=`Survey Year`, y=expenses,  group=team, color=team)) +
   scale_y_continuous(labels = scales::comma, limits = c(0, 30000000)
)

#this label list is huge, but we can see men's basketball sticking out. let's zoom in a bit.
```


#Let's take a look at a more specific selection of sports. The first one was too broad. All of the labels and lines were smushed.
```{r}
#new sports selection, excluding swim and dive, gymnastics, water polo
md_teams_expenses_selects <- expenses%>%
  select(`Institution Name`, `Survey Year`, `Basketball Men's Team Expenses`, `Basketball Women's Team Expenses`, `Baseball Total Expenses`,`Field Hockey Total Expenses`, `Football Total Expenses`, `Lacrosse Men's Team Expenses`, `Lacrosse Women's Team Expenses`, `Soccer Men's Team Expenses`, `Soccer Women's Team Expenses`, `Softball Total Expenses`,  `Volleyball Total Expenses`, `Wrestling Total Expenses`)


#prep for visualization by pivoting
md_teams_expenses_long_selects <-
md_teams_expenses_selects %>% pivot_longer(cols=c("Basketball Men's Team Expenses", "Basketball Women's Team Expenses", "Baseball Total Expenses","Field Hockey Total Expenses", "Football Total Expenses", "Lacrosse Men's Team Expenses", "Lacrosse Women's Team Expenses", "Soccer Men's Team Expenses", "Soccer Women's Team Expenses", "Softball Total Expenses", "Volleyball Total Expenses", "Wrestling Total Expenses"),
                    names_to='team',
                    values_to='expenses')


ggplot() + 
  geom_line(data=md_teams_expenses_long_selects, aes(x=`Survey Year`, y=expenses,  group=team, color=team)) +
  labs(
    title="Maryland spends the vast majority of its athletics money on football") + 
   scale_y_continuous(labels = scales::comma,)

```
This visualization is a lot easier to take in. It makes the breadth of football spending visible. Men's and women's basketball are also huge. 

#Now, let's exclude football.
```{r}
#select list of teams
md_teams_expenses_selects_big_nofb <- expenses%>%
  select(`Institution Name`, `Survey Year`, `Basketball Men's Team Expenses`, `Basketball Women's Team Expenses`, `Baseball Total Expenses`,`Field Hockey Total Expenses`, `Lacrosse Men's Team Expenses`, `Lacrosse Women's Team Expenses`, `Soccer Men's Team Expenses`, `Soccer Women's Team Expenses`, `Softball Total Expenses`,  `Volleyball Total Expenses`)

#pivot to prep for visualization
md_teams_expenses_long_selects_big_nofb <-
md_teams_expenses_selects_big_nofb %>% pivot_longer(cols=c("Basketball Men's Team Expenses", "Basketball Women's Team Expenses", "Baseball Total Expenses","Field Hockey Total Expenses","Lacrosse Men's Team Expenses", "Lacrosse Women's Team Expenses", "Soccer Men's Team Expenses", "Soccer Women's Team Expenses", "Softball Total Expenses", "Volleyball Total Expenses"),
                    names_to='team',
                    values_to='expenses')
#visualize!
ggplot() + 
  geom_line(data=md_teams_expenses_long_selects_big_nofb, aes(x=`Survey Year`, y=expenses,  group=team, color=team)) + 
  labs(
    title="Outside of football, Maryland spends a lot of money on basketball") + 
   scale_y_continuous(labels = scales::comma,)

```
Here, we can see just how big the gap between men's and women's basketball is. We can also see the gap between basketball and all of these other sports.


#Let's look at just basketball.
```{r}

#select columns
md_teams_expenses_selects_bball <- expenses%>%
  select(`Institution Name`, `Survey Year`, `Basketball Men's Team Expenses`, `Basketball Women's Team Expenses`)

#pivot
md_teams_expenses_long_selects_bball <-
md_teams_expenses_selects_bball %>% pivot_longer(cols=c("Basketball Men's Team Expenses", "Basketball Women's Team Expenses"),
                    names_to='team',
                    values_to='expenses')
#visualize
ggplot() + 
  geom_line(data=md_teams_expenses_long_selects_bball, aes(x=`Survey Year`, y=expenses,  group=team, color=team)) +
  labs(
    title="UMD spends more on men's basketball than it does on women's basketball") + 
   scale_y_continuous(labels = scales::comma,)

```

#Let's compare revenues between men's and women's basketball.

```{r}

md_bball_revenues <- revenue%>%
  select(`Institution Name`, `Survey Year`, `Basketball Men's Team Revenue`, `Basketball Women's Team Revenue`)


long_bball_revenues <- md_bball_revenues %>% pivot_longer(cols=c("Basketball Men's Team Revenue", "Basketball Women's Team Revenue"),
                    names_to='team',
                    values_to='revenue')


ggplot() + 
  geom_line(data=long_bball_revenues, aes(x=`Survey Year`, y=revenue,  group=team, color=team)) +
  labs(
    title="Maryland makes much more from men's basketball than women's basketball") + 
   scale_y_continuous(labels = scales::comma,)

```
Hmm, that's a funky looking pattern for men's basketball. It looks like the gap widened before COVID and narrowed afterward. Because of the nature of the pandemic, I feel like we can't quite make any judgement on what's going on with that decline. Still of note, though, women's basketball did trend up. It's interesting that men's basketball is a bit erratic but women's basketball is very steady. 

It would also be interesting to look at recruiting expenses, but it seems the EADA does not break down this data by team, as far as my research has shown.But we do know that Maryland spends more on men's recruiting. I'm sure football is a big contributor to this, but we can't narrow to that given the data that we have. 

#Now, let's look at winning records for men's and women's basketball.
```{r}
#load in men's basketball data
md_mbb <- read.csv("terps_season_records_mens.csv")
md_mbb <- setnames(md_mbb, "WL.", "mens_record", skip_absent=TRUE)


#join men's basketball season record data with revenues
md_mbb_full <- right_join(md_mbb, md_bball_revenues, by=c('Year'='Survey Year'))%>%
  select("Year", "Basketball Men's Team Revenue", "Basketball Women's Team Revenue","mens_record", "Head.Coaches")

#read in men's basketball data
md_wbb <- read.csv("terps_season_records_womens.csv")
md_wbb <- setnames(md_wbb, "WL.", "womens_record", skip_absent=TRUE)

#join on year
md_bb_full <- right_join(md_mbb_full, md_wbb, by=c('Year'='Year'))

#build plot
ggplot() + 
  geom_line(data=md_wbb, aes(x=Year, y=womens_record), color="blue") +
  geom_line(data=md_mbb, aes (x=Year, y=mens_record), color="green") +
  labs(
    title="Maryland women's basketball wins more than men's basketball",
    y="win-percentage",) +
   scale_y_continuous(labels = scales::comma,) +
    scale_x_continuous(limits = c(2003, 2020))

```
Maryland women's basketball, represnted in blue, has has a higher winning percentage than men's basketball since about 2010. Still, Maryland athletics is spending more on men's basketball and making more on men's basketball.

#Returning to our working question: How has the gap between men's and women's sports changed at UMD over the years?

--> When comparing Maryland to the rest of the Big 10, the expenses gap between men's and women's sports is on the smaller side. 


This question is multifaceted. Through equity reports, I plan to explore the following elements over the past ten years:
--> The difference in revenue for men's and women's sports
--> The difference in total expenses (including game-day expenses) for men's and women's sports
--> The difference in expenses for recruiting in men's and women's sports
In conclusion, the discrepancies are pretty clear. Data shows significant discrepancies in the money Maryland athletics has spent on women's vs. men's sports.