---
title: "Maryland Sports Venues Data Exploration"
output: html_document
---

# Exploring Sports Recreation Deserts in Maryland

In this presentation, I set out to explore the role demographics play in understanding the locations of different sports venues in Maryland. I wanted to search for sports recreational deserts -- areas that aren't close to sports venues, and I wanted to determine what demographic patterns look like.

I began with an exploration at the county level and later dug into demographics and venue data at the Census tract level, to see if there were broader trends at play. **Ultimately, I'll show how areas in the eastern side of Maryland lack access to sports recreation venues, and are also among some of the lower income areas in the state.**

In this analysis, I'll dig into a couple of types of analysis:

-   Counting the number of sports venues by county, by sport

-   Plot the relationship between county population and number of county sports venues

-   Mapping demographics at the Census tract level with sports venue locations

-   Exploring Maryland income levels and sports venue locations

-   Mapping areas with and without selected sports venues

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I began by loading a ton of libraries and my Census API Key.

```{r}
library(tidyverse)
library(sf)
library(janitor)
library(broom)
library(ggplot2)
library(gt)
library(leaflet)
library(dplyr)
library(readxl)
library(shinydashboard)
library(rgdal)
library(htmltools)
library(tidycensus)
library(stringr)

census_api_key("0f4adff4206717d4947de6eb32d22e0b484ae6b8", install = TRUE, overwrite = TRUE)
```

**The data**

I then brought in data on sports venues from the state's open data portal. The databases I'm analyzing were last updated in 2014. As far as I can tell, the basketball dataset only includes venues with indoor courts. It also includes some recreation centers (Leonard Hall Recreation Center) and some high schools (Dr. Henry A. Wise, Jr. High School).

The site describes the data, as follows:

"These venues offer opportunities to participate in free and fee-based - organized and pick-up - indoor and outdoor sports and physical fitness related activities in the area of Basketball."

That leaves many questions, and I'm not sure how accurate this is. Throughout this presentation, bear in mind that this dataset might not be complete and it might be out

```{r}
bball <- st_read("basketball/SOCI_Basketball_MDSports.shp")

golf <- st_read("golf/SOCI_Golf_MDSports.shp")

skateboarding <- st_read("skateboarding/SOCI_Skateboarding_MDSports.shp")

soccer <- st_read("soccer/SOCI_Soccer_MDSports.shp")

swimming <- st_read("swimming/SOCI_Swimming_MDSports.shp")

indoor_sports <- st_read("indoor_sports/SOCI_IndoorSports_MDSports.shp")

roller_sports <- st_read("roller_sports/SOCI_RollerSports_MDSports.shp")

baseball <- st_read("baseball/SOCI_Baseball_MDSports.shp")

ice_skating <- st_read("ice_skating/SOCI_IceSkating_MDSports.shp")

equestrian <- st_read("equestrian/SOCI_Equestrian_MDSports.shp")

tennis <- st_read("tennis/SOCI_Tennis_MDSports.shp")

#combine all
sports_combo <- bind_rows(bball, golf, skateboarding, soccer, swimming, indoor_sports, roller_sports, baseball, ice_skating, equestrian, tennis)

#Check for repeats
sports_combo %>% distinct(Title)%>%
  group_by(Title)%>%
  summarise(count = n())%>%
  arrange(desc(count))
#No repeats!

```

**Examining Sports by County**

Next, I began to explore the locations of sports venues at the county level.

```{r}

bball_counties <- bball%>%
  group_by(County)%>%
  summarize(venue_count = n ())%>%
  select(County, venue_count)%>%
  arrange(desc(venue_count))%>%
  mutate(County = str_to_lower(County))


bball_counties %>% gt()%>%
  cols_hide(columns = c(geometry))%>%
  cols_label(
    County = "Jurisdiction",
    venue_count = "Count of Basketball Courts"
  )%>%
  tab_header(
    title = "Frederick County is filled with basketball courts"
  )%>%
   tab_source_note(
    source_note = "Source: MD iMAP Data Catalog | By Rina Torchinsky"
  )

```

Frederick County is coming out on top with basketball courts, with Baltimore and Baltimore County not too far behind. Let's see which county comes up on top for tennis.

```{r}

tennis_counties <- tennis%>%
  group_by(County)%>%
  summarize(venue_count = n ())%>%
  select(County, venue_count)%>%
  arrange(desc(venue_count))%>%
  mutate(County = str_to_lower(County))


tennis_counties %>% gt()%>%
  cols_hide(columns = c(geometry))%>%
  cols_label(
    County = "Jurisdiction",
    venue_count = "Count of Tennis Venues"
  )%>%
  tab_header(
    title = "Baltimore County leads in tennis venues"
  )%>%
  tab_source_note(
    source_note = "Source: MD iMAP Data Catalog | By Rina Torchinsky"
  )

```

Baltimore County is number one for tennis venues. Frederick is also up top. Let's see if Frederick continues its reign.

Let's look at baseball, too.

```{r}

baseball_counties <- baseball%>%
  group_by(County)%>%
  summarize(venue_count = n ())%>%
  select(County, venue_count)%>%
  arrange(desc(venue_count))%>%
  mutate(County = str_to_lower(County))


baseball_counties %>% gt()%>%
  cols_hide(columns = c(geometry))%>%
  cols_label(
    County = "Jurisdiction",
    venue_count = "Count of Baseball Venues"
  )%>%
  tab_header(
    title = "Frederick County leads in baseball venues"
  )%>%
   tab_source_note(
    source_note = "Source: MD iMAP Data Catalog | By Rina Torchinsky"
  )

```

Frederick County is having a moment. The region is up top for baseball venues by a lot. I guess that might make sense given the physical topography of the area. Also, the Frederick Keys come to mind.

Let's check out ice skating.

```{r}

ice_counties <- ice_skating%>%
  group_by(County)%>%
  summarize(venue_count = n ())%>%
  select(County, venue_count)%>%
  arrange(desc(venue_count))%>%
  mutate(County = str_to_lower(County))


ice_counties %>% gt()%>%
  cols_hide(columns = c(geometry))%>%
  cols_label(
    County = "Jurisdiction",
    venue_count = "Count of Ice Skating Venues"
  )%>%
  tab_header(
    title = "Prince George's County leads in ice skating venues"
  )%>%
   tab_source_note(
    source_note = "Source: MD iMAP Data Catalog | By Rina Torchinsky"
  )

```

This is a smaller dataset. Prince George's County tops the list. Baltimore City and Montgomery County aren't far behind. This might hint at a connection to population, but the small sample size is also playing a role.

Up next, soccer.

```{r}

soccer_counties <- soccer%>%
  group_by(County)%>%
  summarize(venue_count = n ())%>%
  select(County, venue_count)%>%
  arrange(desc(venue_count))%>%
  mutate(County = str_to_lower(County))


soccer_counties %>% gt()%>%
  cols_hide(columns = c(geometry))%>%
  cols_label(
    County = "Jurisdiction",
    venue_count = "Count of Soccer Venues"
  )%>%
  tab_header(
    title = "Frederick County leads in soccer spots"
  )%>% 
tab_source_note(
    source_note = "Source: MD iMAP Data Catalog | By Rina Torchinsky"
  )

```

Frederick once again! This is a new showing from Calvert, though.

Let's look at swimming.

```{r}

swim_counties <- swimming%>%
  group_by(County)%>%
  summarize(venue_count = n ())%>%
  select(County, venue_count)%>%
  arrange(desc(venue_count))%>%
  mutate(County = str_to_lower(County))


swim_counties %>% gt()%>%
  cols_hide(columns = c(geometry))%>%
  cols_label(
    County = "Jurisdiction",
    venue_count = "Count of Swimming Venues"
  )%>%
  tab_header(
    title = "Baltimore County leads in swimming venues"
  )%>%
   tab_source_note(
    source_note = "Source: MD iMAP Data Catalog | By Rina Torchinsky"
  )

```

Baltimore County and Baltimore City are up top.

Rather than continuing to move through sports one by one, we'll throw a ton of sports together and see which counties have the most sports venues.

This broad metric includes:

-   basketball

-   golf

-   skateboarding

-   soccer

-   swimming

-   indoor sports

-   roller sports

-   baseball

-   ice_skating

-   equestrian

-   tennis

```{r}

combo_counties <- sports_combo%>%
  group_by(County)%>%
  summarize(venue_count = n ())%>%
  select(County, venue_count)%>%
  arrange(desc(venue_count))%>%
  mutate(County = str_to_lower(County))


combo_counties %>% gt()%>%
  cols_hide(columns = c(geometry))%>%
  cols_label(
    County = "Jurisdiction",
    venue_count = "Number of sports venues"
  )%>%
  tab_header(
    title = "Baltimore County is filled with sports venues"
  )%>%
   tab_source_note(
    source_note = "Source: MD iMAP Data Catalog | By Rina Torchinsky"
  )

```

We're seeing Baltimore County up top with Montgomery and Frederick counties. All of our Frederick repeats make sense now.

Let's take a look at that a bit more visually, too. As a reminder, the combo_counties analysis includes tennis, basketball, golf, baseball, soccer and ice skating, among other sports venues.

```{r}
combo_counties$County = str_to_title(combo_counties$County)

county_bars <- ggplot(data=combo_counties, aes(x=reorder(County, `venue_count`),y=`venue_count`))+
  geom_bar(stat="identity",fill="red")+
    coord_flip() +
    theme_minimal()+
  labs(title="Baltimore County is Loaded with Sports Venues", subtitle="Of all Maryland jurisdictions, this county comes out on top for most selected sports venues.", y="Count of Venues", x="Jurisdiction", "Source: Maryland iMap Data Catalog | By Rina Torchinsky")
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )

county_bars
```

**Diving into Census Data**

Now that we've got a sense of which counties have a lot of sports venues, it's time to pull in Census data to see if we can bring any meaning to why these counties might be coming out on top.

We'll begin by pulling a big dataset pf variables from tidycensus.

```{r}

#I used this to explore variables of interest.
vars <- load_variables(2010, "acs5", cache = TRUE)

```

I chose a couple of metrics of interest:

-   population with less than high school

-   income

-   population

-   Black population

-   white population

```{r}

less_than_hs_county <- get_acs(geography = "county", 
                     variables = "B23006_002", 
                     state = "MD",
                     geometry = TRUE)

md_income_county <- get_acs(geography = "county", 
                     variables = "B19013_001", 
                     state = "MD",
                     geometry = TRUE)

md_pop_county <- get_acs(geography = "county", 
                     variables = "B01003_001", 
                     state = "MD",
                     geometry = TRUE)


md_black_pop_county <- get_acs(geography = "county", 
                     variables = "B02001_003", 
                     state = "MD",
                     geometry = TRUE)


md_white_pop_county <- get_acs(geography = "county", 
                     variables = "B02001_002", 
                     state = "MD",
                     geometry = TRUE)

```

I'll pull in some similar data by Census tract. We'll explore this later.

```{r}

less_than_hs_tract <- get_acs(geography = "tract", 
                     variables = "B23006_002", 
                     state = "MD",
                     geometry = TRUE)

md_income_tract <- get_acs(geography = "tract", 
                     variables = "B19013_001", 
                     state = "MD",
                     geometry = TRUE)

md_black_pop_tract <- get_acs(geography = "tract", 
              variables = "B02001_003", 
              state = "MD", 
              geometry = TRUE)
```

Bearing in mind the populous counties that had the most venues, I wanted to explore which counties have the most basketball venues per person.

```{r}
md_race_pct <- get_acs(geography = "county", 
              variables = c(black_pop = "B02001_003", total_pop = "B01003_001", white_pop = "B02001_002"), 
              state = "MD", 
              year = 2010)%>%
              select(GEOID, NAME, variable, estimate)

md_race_wider <- md_race_pct%>%
  pivot_wider(names_from = variable, values_from = estimate)%>%
  mutate(pct_black = black_pop/total_pop)%>%
  mutate(pct_white = white_pop/total_pop)%>%
  mutate(county_clean = str_to_lower(NAME))

md_pop_clean <-md_race_wider%>%
  separate(county_clean, c("County", "state"), sep = " c", extra = "merge")

md_bball_pop_county <- inner_join(md_pop_clean, bball_counties)%>%
  select(County, venue_count, total_pop, black_pop, pct_black, white_pop, pct_white)%>%
  mutate(per_pop = venue_count/total_pop)%>%
  arrange(desc(per_pop))

md_bball_pop_county

```

Here, we calculated two new metrics: the percent of a jurisdiction's population that is Black and the number of basketball venues per jurisdiction. It was interesting to see Somerset, Kent, Frederick, Garrett and Caroline up there.

These counties have smaller populations, so these numbers are definitely inflated. Garrett has a single basketball venue listed, but is in the top. These numbers, of course, are not telling the full story. Frederick's repeated appearance toward the top could be notable, though.

Let's dig back into the sports combination dataset, and see if there's anything of interest there -- and to see if we see the same counties.

```{r}

combo_counties$County = str_to_lower(combo_counties$County)

sports_combo_county <- inner_join(md_pop_clean, combo_counties)%>%
  select(County, venue_count, total_pop, black_pop, pct_black, white_pop, pct_white)%>%
  mutate(per_pop = venue_count/total_pop)%>%
  arrange(desc(per_pop))

sports_combo_county

```

Somerset, Kent, Worcester and Queen Anne's are leaders here. They also have smaller populations.

We can plot it, too. When we do, we see an interesting line here. More venues, more population. This is not shocking.

```{r}
ggplot() + geom_point(data=sports_combo_county, aes(x=venue_count, y=total_pop)) +
  geom_smooth(data=sports_combo_county, aes(x=venue_count, y=total_pop), method="lm")+
  labs(title="More People, More Sports Venues", subtitle= "Maryland areas with bigger populations have more sports venues.", "Source: Maryland iMap Data Catalog, U.S. Census | By Rina Torchinsky", x="Count of Sports Venues", y="Population")
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 5), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
```

This isn't a wild surprise, but we can see that more higher county population also means a higher count of sports venues. Population is playing a huge role, so it's important we zero in tighter than the county level. Since this project was aimed at searching for sports recreational deserts, I want to begin to explore the geographic location of the sports venues.

**Mapping Selected Sports**

Let's start by mapping a collection of sports venues to get a better sense of what's going on at the local level.

```{r}

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    data=bball,
    color="orange",
    radius=2,
    group="Basketball")%>%
  addCircleMarkers(
    data=golf,
    color="green",
    radius=2,
    group="Golf")%>%
  addCircleMarkers(
    data=skateboarding,
    color="purple",
    radius=2,
    group="Skateboarding")%>%
  addCircleMarkers(
    data=soccer,
    color="black",
    radius=2,
    group="Soccer")%>%
  addCircleMarkers(
    data=swimming,
    color="blue",
    radius=2,
    group="Swimming")%>%
  addLayersControl(
    overlayGroups = c("Golf", "Basketball", "Soccer", "Swimming", "Skateboarding"),
    options = layersControlOptions(collapsed = FALSE)
  )
  
```

There's definitely a concentration around the D.C. Metropolitan Area and Baltimore. Areas around the Maryland-Delaware border are lacking sports venues.

Let's see what happens if we load in some demographic information, and keep the sports. It's going to be a bit chaotic at first glance, but if we interact with it and toggle some of the overlay features, we can check out which demographics seem to be tightly connected.

```{r}
md_shp <- st_read("https://raw.githubusercontent.com/frankrowe/maryland-geojson/master/maryland.geojson")

pal5 <- colorQuantile(palette = "BuPu", domain = md_income_tract$estimate, n = 10)

pal6 <- colorQuantile(palette = "BuPu", domain = md_black_pop_tract$estimate, n = 10)

numPal <- colorNumeric('viridis', md_income_tract$estimate)

quantPal <- colorQuantile('viridis', md_income_tract$estimate, n = 5)

pal <- colorQuantile(
  palette = "viridis",
  domain = md_income_tract$estimate, n=8, na.color = NA)

mypal <- colorQuantile(
  palette = "viridis",
  domain = md_black_pop_tract$estimate, n=8, na.color = NA)

thepal <- colorQuantile(
  palette = "viridis",
  domain = less_than_hs_tract$estimate, n=5, na.color = NA)

md_map_all <- 
  leaflet()%>%
  addProviderTiles(providers$CartoDB.Positron)%>%
  addPolylines(
    data=md_shp,
    weight=2,
    color="gray"
  )%>%
  addPolygons(data = md_income_tract, group = "Income by Tract", color = ~ pal(estimate), stroke = FALSE)%>%
  addPolygons(data = md_black_pop_tract, group = "Black Population by Tract", color = ~ mypal(estimate), stroke = FALSE)%>%
  addPolygons(data = less_than_hs_tract, group = "Less Than High School by Tract", color = ~ thepal(estimate), stroke = FALSE)%>%
  addLegend("bottomleft", pal = pal, values = md_income_tract$estimate, title = "Income Quantiles")%>%
  addLegend("bottomleft", pal = mypal, values = md_black_pop_tract$estimate, title = "Black Population Quantiles")%>%
   addLegend("bottomright", pal = mypal, values = less_than_hs_tract$estimate, title = "Less Than High School Quantiles")%>%
  addCircleMarkers(
    data=bball,
    color="orange",
    radius=2,
    group="Basketball")%>%
  addCircleMarkers(
    data=golf,
    color="green",
    radius=2,
    group="Golf")%>%
  addCircleMarkers(
    data=skateboarding,
    color="purple",
    radius=2,
    group="Skateboarding")%>%
  addCircleMarkers(
    data=soccer,
    color="black",
    radius=2,
    group="Soccer")%>%
  addCircleMarkers(
    data=swimming,
    color="blue",
    radius=2,
    group="Swimming")%>%
   addLayersControl(
  overlayGroups = c("Golf", "Basketball", "Soccer", "Swimming", "Skateboarding"),
     baseGroups = c("Income by Tract", "Black Population by Tract", "Less Than High School Quantiles"),
  options = layersControlOptions(collapsed = FALSE)) 
  
md_map_all
```

It helps to make this full screen to toggle through all of it.

Income seems to be the most closely related. When you toggle to income by tract, you can see the way sports venues are concentrated in the yellow areas, which are the highest income areas. Many areas on the eastern part of Maryland are purple and blue, indicating lower-income. These areas also largely lacked sports venues.

The other demographics I explored -- education and Black population -- weren't as closely connected to the sports venues. For example, if you scroll to parts of Baltimore with high Black populations , some areas still have sports venues. High School education levels also did not appear to be closely connected. Income, stands out, though, when you begin to explore the income overlay paired with the

Knowing that income was closely correlated, I created a clean map without the clutter of the added legends and base layers. Here, we look only at income, just to get a fresher, more presentable view.

```{r}
md_map_income <- 
  leaflet()%>%
  addProviderTiles(providers$CartoDB.Positron)%>%
  addPolylines(
    data=md_shp,
    weight=2,
    color="gray"
  )%>%
  addPolygons(data = md_income_tract, group = "Income by Tract", color = ~ pal(estimate), stroke = FALSE)%>%
  addLegend("bottomleft", pal = pal, values = md_income_tract$estimate, title = "Income Quantiles")%>%
  addCircleMarkers(
    data=bball,
    color="orange",
    radius=2,
    group="Basketball")%>%
  addCircleMarkers(
    data=golf,
    color="green",
    radius=2,
    group="Golf")%>%
  addCircleMarkers(
    data=skateboarding,
    color="purple",
    radius=2,
    group="Skateboarding")%>%
  addCircleMarkers(
    data=soccer,
    color="black",
    radius=2,
    group="Soccer")%>%
  addCircleMarkers(
    data=swimming,
    color="blue",
    radius=2,
    group="Swimming")%>%
   addLayersControl(
  overlayGroups = c("Golf", "Basketball", "Soccer", "Swimming", "Skateboarding"),
  options = layersControlOptions(collapsed = FALSE)) 
  
md_map_income

```

Noting our interest in the eastern part of Maryland, we notice a cluster of golf courses in Ocean City, Maryland, which is also shown to be on the lower end of the income quantiles. There's also a cluster of sports venues not far from Salisbury, which is also on the lower end of the income quantiles. But south of Cambridge, right by the Chesapeake, there's almost nothing.

**Diving Deeper Into Income Levels**

Next, we'll get a sense of income levels. We'll create five income buckets and see where sports are concentrated.

First, we'll look at golf.

```{r}

golf_st <- st_transform(golf, 4269)

md_income_tract_binned <- md_income_tract %>% mutate(new_bin = ntile(estimate, n=5))

md_income_tract_golf_bins <- golf_st %>%
  st_join(md_income_tract_binned)%>%
  group_by(new_bin)%>%
  summarize(count = n())


golf_income <-ggplot(data=md_income_tract_golf_bins, aes(x=new_bin, y=count))+
  geom_bar(stat="identity",fill="#5a9e4b")+
  labs(title="Golf courses are concentrated in wealthy Maryland areas", caption="Source: Maryland iMap Data Catalog, U.S. Census | By Rina Torchinsky", x="Income Buckets", y="Sports Venues")  + theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )

golf_income

```

This isn't a wild surprise. There are more golf courses in wealthier Census tracts. Golf can be an expensive sport. Let's check out basketball -- a personal favorite.

```{r}

bball_st <- st_transform(bball, 4269)

md_income_tract_bball_bins <- bball_st %>%
  st_join(md_income_tract_binned)%>%
  group_by(new_bin)%>%
  summarize(count = n())

bball_income <-ggplot(data=md_income_tract_bball_bins, aes(x=new_bin, y=count))+
  geom_bar(stat="identity",fill="orange")+
  labs(title="Basketball venues are in high and low income parts of Maryland", caption="Source: Maryland iMap Data Catalog, U.S. Census | By Rina Torchinsky", x="Income Buckets", y="Sports Venus")  + theme_minimal() + 
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )

bball_income


```

This is interesting. Basketball exists in a lot of lower income areas, but also in a lot of very high income areas. This follows a much different trend than golf.

```{r}
soccer_st <- st_transform(soccer, 4269)

md_income_tract_soccer_bins <- soccer_st %>%
  st_join(md_income_tract_binned)%>%
  group_by(new_bin)%>%
  summarize(count = n())

soccer_income <-ggplot(data=md_income_tract_soccer_bins, aes(x=new_bin, y=count))+
  geom_bar(stat="identity",fill="purple")+
  labs(title="Soccer spots are split between high and low income Maryland areas", x="Income Buckets", y="Sports Venues", caption="Source: Maryland iMap Data Catalog, U.S. Census | By Rina Torchinsky")  + theme_minimal() + 
  theme(
    plot.title = element_text(size = 13, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
soccer_income


```

Although soccer venues are showing up in lower-income areas, the higher-income areas are packed. The lack of soccer venues in the middle of the income range is also notable, and seems to somewhat echo what we say with basketball.

```{r}
swimming_st <- st_transform(swimming, 4269)

md_income_tract_swim_bins <- swimming_st %>%
  st_join(md_income_tract_binned)%>%
  group_by(new_bin)%>%
  summarize(count = n())

md_income_tract_swim_bins


swim_income <-ggplot(data=md_income_tract_swim_bins, aes(x=new_bin, y=count))+
  geom_bar(stat="identity",fill="darkblue")+
  labs(title="Swimming pools cover high and low income Maryland areas", x="Income Buckets", y="Sports Venues", caption="Source: Maryland iMap Data Catalog, U.S. Census | By Rina Torchinsky")  + theme_minimal() + 
  theme(
    plot.title = element_text(size = 13.5, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
swim_income
```

The lowest income bucket has the highest number of swimming venues, but when we begin to take a more holistic view, it's easy to see that the pools are filling the high two income buckets, in general.

**Pulling Sports Together**

Let's begin to pull in a collection of sports.

```{r}

combo_st <- st_transform(sports_combo, 4269)

combo_bins <- combo_st %>%
  st_join(md_income_tract_binned)%>%
  group_by(new_bin)%>%
  summarize(count = n())

combo_income <-ggplot(data=combo_bins, aes(x=new_bin, y=count))+
  geom_bar(stat="identity",fill="lightblue")+
labs(title="Higher income Maryland areas are packed with sports venues", subtitle="Lower income areas do have sports venues, but top income areas have much more.", x="Income Buckets", y="Sports Venues", caption="Source: Maryland iMap Data Catalog, U.S. Census | By Rina Torchinsky")  + theme_minimal() + 
  theme(
    plot.title = element_text(size = 13.5, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
combo_income


```

We're going to look at tracts with and tracts without sports venues. Of note, there are some NAs that we're weeding out for the purpose of mapping.

```{r}

tract_income_combo <- combo_st %>%
  st_join(md_income_tract)

#count of number of venues per tract
combo_analysis <- tract_income_combo %>%
  group_by(GEOID)%>%
  drop_na(GEOID)%>%
  summarize(count= n())%>%
  arrange(desc(count))

#check for distinctions and count 
all_tracts <- md_income_tract%>%distinct(GEOID)

#grab tracts without sports. These are ones.
sportsless_tracts <- anti_join(all_tracts, combo_analysis)%>%
  mutate(col = 0)

#create mappable dataframe of tracts with venues.
df <- left_join(sportsless_tracts, md_income_tract)

#create mappable dataframe of tracts without venues.
df_w <- left_join(md_income_tract, sportsless_tracts)%>%
  replace(is.na(.), 1)%>%
  filter(col == 1)

pal30 <- colorNumeric(palette = "yellow", domain = df$col)

pal20 <- colorNumeric(palette = "red", domain = df$col)

df <- st_as_sf(df)

```

```{r}

counties <- st_read("counties.geojson")

  
leaflet()%>%
  addProviderTiles(providers$CartoDB.Positron)%>%
  addPolygons(data = df_w, group= "With Sports", color = ~ pal30(col), stroke = FALSE)%>%
  addPolygons(data = df, group= "Without Sports", color = ~ pal20(col), stroke = FALSE)%>%
  addPolygons(data = md_income_tract, group = "Income Levels", color = ~ pal(estimate), stroke = FALSE)%>%
  addPolylines(
    data=md_shp,
    weight=2,
    color="gray"
  )%>%
  addPolylines(
    data=counties,
    weight=2,
    color="gray"
  )%>%
  addLegend("bottomleft", pal = pal, values = md_income_tract$estimate, title = "Income Quantiles")%>%
  addLayersControl(
     overlayGroups = c("Without Sports", "With Sports", "Income Levels"),
  options = layersControlOptions(collapsed = FALSE)) 
  
```

There are a couple of ways that I interacted with this map to identify areas that lack sports venues. I toggled to just "Without Sports" to identify areas that lack sports venues. Then, I switched off that layer and turned on the income layer, so I could determine if areas without sports were also lower income.

Nearly all of Dorchester County does not have any sports venues. The entire county is also predominantly lower income. Caroline County also lacks sports venues outside of the area around Denton. If you scroll to the county, you can see that there are some sports venues in Caroline County around Denton, but once you approach Federalsburg, which is lower-income, there are no longer any venues in the area. Greensboro, which is in the northern part of Caroline County is lower income and has no sports venues.

Much of Allegany County also lacks sports venues, as does the western portion of Washington County, which is a lower-income region. The area south of Oakland, which is in some of the lowest income quantiles, has no sports venues.

Parts of Prince George's County that hug Washington, D.C., including Chillum and Mount Rainier are lower income and lack sports venues. On the contrary, upper parts of Montgomery County, including Poolesville and Barnesville, are wealthier and also have sports venues. In addition, the area north of Columbia in Howard County is both higher income and has sports venues.\
\
**Conclusion and Next Steps**

Sports recreational deserts exist in Maryland, and they do exist in the lower income areas identified above. The analysis shows that wealthier areas have sports venues. Next, it would be interesting to explore change over time, as well as what sort of decision making goes into the construction of new sports venues. Cleaner, newer, more expansive data, would also be important to making stronger points about these sports recreational deserts.

It would also be interesting to explore the next level of "why it matters," by exploring high school sports participation or similar data. I would be interested to identify the impact of sports recreational deserts -- the athletes they keep away and the athletes who succeeds despite them.
